Relative parabolic factorizations of Chevalley groups

### 2015-08-24 paper.tex
Slightly reorganized exposition

\DeclareMathOperator{\K}{K}
\DeclareMathOperator{\G}{G}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\SL}{SL}
\DeclareMathOperator{\Sp}{Sp}
\DeclareMathOperator{\St}{St}
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\EP}{EP}
\DeclareMathOperator{\Par}{P}
\DeclareMathOperator{\B}{B}
\DeclareMathOperator{\Hh}{H}
\DeclareMathOperator{\U}{U}
\DeclareMathOperator{\X}{X}
\DeclareMathOperator{\Z}{Z}
\DeclareMathOperator{\M}{M}
\DeclareMathOperator{\SR}{SR}
\DeclareMathOperator{\sr}{sr}
\DeclareMathOperator{\shape}{shape}
\DeclareMathOperator{\Rad}{Rad}
\DeclareMathOperator{\Max}{Max}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Spin}{Spin}
\DeclareMathOperator{\Stab}{Stab}
\DeclareMathOperator{\ASR}{ASR}
\DeclareMathOperator{\asr}{asr}
%TODO: Deprecate Um -- Use Ums/Umd whenever possible
\DeclareMathOperator{\Um}{Um}
\DeclareMathOperator{\Ums}{Ums}
\DeclareMathOperator{\Umd}{Umd}
\DeclareMathOperator{\rk}{rk}
\newcommand{\rA}{\mathsf{A}}
\newcommand{\rB}{\mathsf{B}}
\newcommand{\rC}{\mathsf{C}} 
\newcommand{\rD}{\mathsf{D}} 
\newcommand{\rE}{\mathsf{E}}

@@

\title{Relative parabolic factorizations of Chevalley groups}
\keywords {Chevalley groups, relative subgroups, stability for $\K_1$. {\em Mathematical Subject Classification (2010):} 20G40, 20G35, 19B14}
\author {Sergey Sinchuk, Andrei Smolensky}
\email {sinchukss {\it at} yandex.ru,\ andrei.smolensky {\it at} gmail.com}
\date {\today}

@@

On the other hand, much less is known about the width of relative groups $\G(\Phi, R, I)$. 
It is a classical result of J.~Tits and L.~Vaserstein that the relative elementary group $\E(\Phi, R, I)$ is generated by elements $z_\alpha(s,\xi)=x_{\alpha}(s)^{x_{-\alpha}(\xi)}$, see Theorem~\ref{theorem:Tits-Vaserstein} below.
Denote by $\mathcal{Z}$ the set of all $z_\alpha(s, \xi)$, $s\in I$, $\xi \in R$. 
Our first application of parabolic factorizations is the following estimate of the width of $\G(\Phi, R, I)$ in terms of $\mathcal{Z}$.

\begin{thm} Let $\Phi$ be a root system, and let $I$ be an ideal of a commutative ring $R$.
\begin{enumerate}
\item If $\sr(I) = 1$, then $W(\E(\Phi,R,I),\mathcal{Z})\leqslant 3|\Phi^+|+2\rk\Phi-1$;
\item If $R = \mathbb{Z}[\sfrac{1}{p}]$ for some prime $p$, and $\Phi$ is classical, then under the assumption of the Generalized Riemann Hypothesis $W(\E(\Phi,R,I),\mathcal{Z})\leqslant ???$;
\item If $\mathcal{O}_S$ is a Dedekind ring of arithmetic type in some global field $k$ having a real embedding and $\rk\Phi\geqslant2$, then $W(\G(\Phi,\mathcal{O}_S,I),\mathcal{Z})$ is finite.
\end{enumerate}
\end{thm}
\begin{proof}
%In the proof of the above theorem we use Bass---Kolster decomposition to reduce to the case $\Phi=\rA_1$ which then can be settled directly.
%Notice that the particular case $I=R$ of the above theorem was demonstrated in~\cite{VavSmSuUnitrEng} (see also \cite{VseUnitrZ1p}).
...
\end{proof}
The particular case $\Phi = \rA_\ell$, $R = \mathbb{Z}$ of the above theorem was considered by U.~Hadad in~\cite[Theorem~1.6]{Ha12}.

@@

\subsection{Root systems}
Let $\Pi=\{\alpha_1, \ldots, \alpha_l\}$ be a basis of $\Phi$. Denote by $\Phi^+$ and $\Phi^-$ subsets of positive and negative roots with respect to $\Pi$.
Denote by $m_i(\alpha)$ the coefficient of $\beta$ in the expansion of $\alpha$ in $\Pi$, i.\,e. $\alpha = \sum_{i=1}^n m_i(\alpha) \alpha_i$.

@@

We denote by $\E(\Phi, R, I)$ the normal closure of $\E(\Phi, I)$ in $\E(\Phi, R)$, in other words $\E(\Phi, R, I) = \E(\Phi, I)^{\E(\Phi, R)}$.
Denote by $z_\alpha(s, \xi)$ the element $x_\alpha(s)^{x\ssub{-\alpha}(\xi)}$.
The following result of L.~Vaserstein and J.~Tits describes a set of generators of $\E(\Phi, R, I)$ (see e.\,g.~\cite[Theorem~2]{Va86}).

\begin{thm}\label{theorem:Tits-Vaserstein} Let $\Phi$ be an irreducible root system of rank $\geq 2$, then the relative elementary subgroup $\E(\Phi, R, I)$
is generated by the elements $z_\alpha(s, \xi)$, $\alpha\in \Phi$, $s\in I$, $\xi\in R$.
\end{thm}

In fact, one can choose a more economical set of generators for $\E(\Phi, R, I)$ (see~\cite[Theorem~3.4]{S}).
\begin{thm}\label{theorem:Stepanov}
Let $\Phi$ be a reduced irreducible root system of rank $\geq 2$ and let $S$ be any parabolic subset of roots of $\Phi$ with the special part $\Sigma_S$.
Then $\E(\Phi, R, I)$ is generated by the union of the following two families of elements:
\begin{itemize}
\item $x_{\alpha}(s)$, where $s\in I$, $\alpha\in\Phi$;
\item $z_\alpha(s,\xi)$, where $s\in I$, $\xi\in R$, $\alpha\in\Sigma_S$.
\end{itemize}
\end{thm}

@@

\section{Applications}\label{sec:applications}
\subsection{Bounded generation}\label{sec:boundgen}
\input{congrsubgr-boundgen}
%\subsection{Subsystem factorizations}\label{sec:subsysfact}
%\input{subsystem-factorizations}

>>>>

\DeclareMathOperator{\K}{K}
\DeclareMathOperator{\G}{G}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\SL}{SL}
\DeclareMathOperator{\Sp}{Sp}
\DeclareMathOperator{\St}{St}
\DeclareMathOperator{\E}{E}
\DeclareMathOperator{\EP}{EP}
\DeclareMathOperator{\Par}{P}
\DeclareMathOperator{\B}{B}
\DeclareMathOperator{\Hh}{H}
\DeclareMathOperator{\U}{U}
\DeclareMathOperator{\X}{X}
\DeclareMathOperator{\Z}{Z}
\DeclareMathOperator{\M}{M}
\DeclareMathOperator{\SR}{SR}
\DeclareMathOperator{\sr}{sr}
\DeclareMathOperator{\shape}{shape}
\DeclareMathOperator{\Rad}{Rad}
\DeclareMathOperator{\Max}{Max}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Spin}{Spin}
\DeclareMathOperator{\Stab}{Stab}
\DeclareMathOperator{\ASR}{ASR}
\DeclareMathOperator{\asr}{asr}
\DeclareMathOperator{\Ums}{Ums}
\DeclareMathOperator{\Umd}{Umd}
\DeclareMathOperator{\rk}{rk}
\newcommand{\rA}{\mathsf{A}}
\newcommand{\rB}{\mathsf{B}}
\newcommand{\rC}{\mathsf{C}} 
\newcommand{\rD}{\mathsf{D}} 
\newcommand{\rE}{\mathsf{E}}

@@

\title [Parabolic factorizations of congruence subgroups ] {Parabolic factorizations and width of congruence subgroups in Chevalley groups}
\keywords {Chevalley groups, relative subgroups, stability for $\K_1$. {\em Mathematical Subject Classification (2010):} 20G40, 20G35, 19B14}
\author {Sergey Sinchuk, Andrei Smolensky}
\email {sinchukss {\it at} yandex.ru,\ andrei.smolensky {\it at} gmail.com}
\date {\today}

@@

On the other hand, much less is known about the width of relative groups $\G(\Phi, R, I)$. 
Set $z_\alpha(s, \xi) = x_{\alpha}(s)^{x_{-\alpha}(\xi)}$, $s\in I$, $\xi\in R$.
It is a classical result of J.~Tits and L.~Vaserstein (see e.\,g. Theorem~\cite[Theorem~2]{Va86}) that the relative elementary group $\E(\Phi, R, I)$ is generated by
the set of elements $\mathcal{Z} = \{ z_\alpha(s,\xi) \mid \alpha \in \Phi,\ s\in I,\ \xi \in R \}.$
Recently, A.~Stepanov has shown that, in fact, a more economical set of generators can be chosen for $\E(\Phi, R, I)$ (see~\cite[Theorem~3.4]{S}).
\begin{thm}
Let $\Phi$ be a reduced irreducible root system of rank $\geq 2$ and let $S$ be any parabolic subset of roots of $\Phi$ with the special part $\Sigma_S$ (see section~\ref{sec:rootsys}).
Denote by $\mathcal{Z}(\Sigma_S)$ the subset of $\mathcal{Z}$ consisting of the following two families of elements:
\begin{itemize}
\item $x_{\alpha}(s)$, where $s\in I$, $\alpha\in\Phi$;
\item $z_\alpha(s,\xi)$, where $s\in I$, $\xi\in R$, $\alpha\in\Sigma_S$.
\end{itemize}
Then $\E(\Phi, R, I)$ is generated (as an abstract group) by $\mathcal{Z}(\Sigma_S)$.
\end{thm}

Our first application of parabolic factorizations is the following theorem which gives an estimate of the width of $\G(\Phi, R, I)$ in terms $\mathcal{Z}$.
\begin{thm} Let $\Phi$ be a root system, and let $I$ be an ideal of a commutative ring $R$.
\begin{enumerate}
\item If $\sr(I) = 1$, then $W(\E(\Phi,R,I),\mathcal{Z})\leqslant 3|\Phi^+|+2\rk\Phi-1$;
\item If $R = \mathbb{Z}[\sfrac{1}{p}]$ for some prime $p$, and $\Phi$ is classical, then under the assumption of the Generalized Riemann Hypothesis $W(\E(\Phi,R,I), \mathcal{Z}(\Sigma_\ell))\leqslant ???$;
\item If $\mathcal{O}_S$ is a Dedekind ring of arithmetic type in some global field $k$ having a real embedding and $\rk\Phi\geqslant2$, then $W(\G(\Phi,\mathcal{O}_S,I), \mathcal{Z}(\Sigma_\ell))$ is finite.
\end{enumerate}
\end{thm}
\begin{proof}

%In the proof of the above theorem we use Bass---Kolster decomposition to reduce to the case $\Phi=\rA_1$ which then can be settled directly.
%Notice that the particular case $I=R$ of the above theorem was demonstrated in~\cite{VavSmSuUnitrEng} (see also \cite{VseUnitrZ1p}).
...
\end{proof}
The particular case $\Phi = \rA_\ell$, $R = I = \mathbb{Z}$ of the above theorem is similar to a recent theorem of U.~Hadad (see ~\cite[Theorem~1.6]{Ha12}).

@@

\subsection{Root systems}\label{sec:rootsys}
Let $\Pi=\{\alpha_1, \ldots, \alpha_l\}$ be a basis of $\Phi$. Denote by $\Phi^+$ and $\Phi^-$ subsets of positive and negative roots with respect to $\Pi$.
Denote by $m_i(\alpha)$ the coefficient of $\beta$ in the expansion of $\alpha$ in $\Pi$, i.\,e. $\alpha = \sum_{i=1}^n m_i(\alpha) \alpha_i$.

@@

We denote by $\E(\Phi, R, I)$ the normal closure of $\E(\Phi, I)$ in $\E(\Phi, R)$, in other words $\E(\Phi, R, I) = \E(\Phi, I)^{\E(\Phi, R)}$.
Denote by $z_\alpha(s, \xi)$ the element $x_\alpha(s)^{x\ssub{-\alpha}(\xi)}$.

%The following result of L.~Vaserstein and J.~Tits describes a set of generators of $\E(\Phi, R, I)$ (see e.\,g.~\cite[Theorem~2]{Va86}).
%\begin{thm}\label{theorem:Tits-Vaserstein} Let $\Phi$ be an irreducible root system of rank $\geq 2$, then the relative elementary subgroup $\E(\Phi, R, I)$
%is generated by the elements $z_\alpha(s, \xi)$, $\alpha\in \Phi$, $s\in I$, $\xi\in R$. \end{thm}

%In fact, one can choose a more economical set of generators for $\E(\Phi, R, I)$ (see~\cite[Theorem~3.4]{S}).
%\begin{thm}\label{theorem:Stepanov}
%Let $\Phi$ be a reduced irreducible root system of rank $\geq 2$ and let $S$ be any parabolic subset of roots of $\Phi$ with the special part $\Sigma_S$.
%Then $\E(\Phi, R, I)$ is generated by the union of the following two families of elements:
%\begin{itemize}
%\item $x_{\alpha}(s)$, where $s\in I$, $\alpha\in\Phi$;
%\item $z_\alpha(s,\xi)$, where $s\in I$, $\xi\in R$, $\alpha\in\Sigma_S$.
%\end{itemize}
%\end{thm}

@@

%\section{Applications} %\label{sec:applications}
\section{Bounded generation}\label{sec:boundgen}
\input{congrsubgr-boundgen}
%\subsection{Subsystem factorizations}\label{sec:subsysfact}
%\input{subsystem-factorizations}
